# knitr document van Steensel lab

# Gene repression in LADs
## Christ Leemans, 03-11-2016 - to date

## Introduction
Generally speaking, genes inside lamina associated domains are not or very lowly expressed. These genes can either be actively repressed by their DNA context (e.g. heterochromatin, lamina association), or simply be inactive (because essential factors for expression are missing?). Yet another group of genes seem to evade gene repression in the context of lamina associated domains. In this report I would like to investigate what defines these 3 groups of genes and how they compare to genes outside of lamina associated domains. Previously we've seen through motif analysis that in general in escaper promoters, there is a higher affininity predicted for ETS-factors and certain bZIPs (motifs with CREB-like binding mode). This is further supported by expression correlation of these factors. I previously focussed on proteins for which there is a Jaspar entry. In this report I would like to focus on expression correlation with the complete set of ETS and bZIP factors. Since the binding motifs of these factors are really similar and this might help us narrow down on which factors are most important. For bZIP factors I am also interested if transcription factors with a predominately CREB-like binding mode (TGACGTCA) are more associated than bZips with JUN-like binding mode (TGANTCA).

```{r}
library(reshape2)
library(rtracklayer)
library(ggplot2)
library(DESeq2)
library(gridExtra)
library(plyr)
library(scales)
library(Matrix)
library(biomaRt)
library(grid)
library(stringr)
library(glmnet)
library(rtfbs)
library(BSgenome)
library(BSgenome.Hsapiens.UCSC.hg19)
library(doMC)
library(ggdendro)
registerDoMC(cores=10)
load('../raw_data/biomart.rdata')

## There was one promoter that was wrongly annotated
bm_p[bm_p$ensembl_transcript_id=='ENST00000357491','ensembl_gene_id' ] = 'ENSG00000196350'



## Did not want to go through the hassle of install complete TFBStools package, since it was
## complaining abount dependencies. And I only needed JASPAR motif parsing functionality.
## So I copied the function from TFBSTools.
## <<< copied from TFBSTools >>>
### -----------------------------------------------------------------
### readJASPARMatrix: read the jaspar format PFM in txt file
### "individual" format:
### >MA0001.1 AGL3
### A  [ 0  3 79 40 66 48 65 11 65  0 ]
### C  [94 75  4  3  1  2  5  2  3  3 ]
### G  [ 1  0  3  4  1  0  5  3 28 88 ]
### T  [ 2 19 11 50 29 47 22 81  1  6 ]
### "all" format: multiple "individual" matrices and seperated with a blank line
### Exported
.processJASPARText <- function(text){
  ID <- sub("^>", "", strsplit(text[1], "\t")[[1]][1])
  name <- strsplit(text[1], "\t")[[1]][2]
  if(!identical(substr(text[2:5], 1, 1), DNA_BASES)){
    stop("The second to fifth lines of the file must start with",
         "`A`, `C`, `G`, `T`, respectively.")
  }
  profileMatrix <- do.call(rbind, strsplit(sub(" *]$", "", 
                                               sub("^(A|C|G|T)  \\[ *", "",
                                                   text[2:5])), " +"))
  mode(profileMatrix) <- "integer"
  rownames(profileMatrix) <- DNA_BASES
  ## changed the following part:
  # ans <- PFMatrix(ID=ID, name=name, profileMatrix=profileMatrix)
  pwm = log(t(profileMatrix)/colSums(profileMatrix))
  ans <- list(ID=ID, name=name, profileMatrix=pwm)
}

readJASPARMatrix <- function(fn, type=c("individual", "all")){
  type <- match.arg(type)
  text <- readLines(fn)
  if(type == "individual"){
    if(length(text) != 5L){
      stop("The `individual` format is supposed to have 5 lines!")
    }
    ans <- .processJASPARText(text)
  }else{
    if(length(text) %% 6 != 0L){
      stop("The `all` format is supposed to have a number of lines",
           "mutipled by 6!")
    }
    text2 <- split(text, rep(1:(length(text)/6), rep(6, length(text)/6)))
    ans <- lapply(text2, .processJASPARText)
    # ans <- do.call(PFMatrixList, ans)
  }
  return(ans)
}
## <<< end of copie >>>>

## get a table with matching sets
## table = complete table to take matching sets from
## class_col = column name of class of interest
## class = name of class to match the set on
## order_on = column name to order on
matchSet <- function(table, class_col, class, order_on){
  o_vec = order(table[,order_on])
  o_table = table[o_vec, ]
  setA = which(o_table[,class_col]==class)
  setB = c(setA + 1, setA -1)
  ## can also return o_table[unique(c(setA, setB)), ]
  ## but this way order is perserved.
  i_vec = o_vec[unique(c(setA, setB))]
  return(table[i_vec[order(i_vec)], ])
}

```



# MAIN

### SuRE data
Previously, Bas defined three different groups of promoters in LADs. The "escaper", "repressed" and "inactive" group. The "escaper" group contains promoters that are expressed in their endogenous LAD environment. The "repressed" group is inactive in the endogenous setting, but they do show autonomous promoter activity in SuRE data. The 'inactive' group shows neither endogenous expression nor autonomous promoter activity in SuRE data.

```{r, fig.width=10, fig.height=8}
load("../raw_data/gencode.sure.160617.rda")
Prom<-gencode.sure.160617; rm(gencode.sure.160617) #simpler name
#first re-calculate pseudocounts without jitter
P<-Prom[,c(1:8,23,20,26,28, 27)] #SuRE, GRO-cap, CAGE and LAD columns only
names(P)[9:13]<-c("SuRE", "GROcap", "CAGE", "LAD", 'tissues_expressed')


## for promoters and gene expression let's convert promoter transcript id's to gene id's
P$ensembl_transcript_id = do.call(rbind, strsplit(P$name, split='[.]'))[,1]

nrow(P) #orriginal number of rows
bm_match = match(P$ensembl_transcript_id, bm_p$ensembl_transcript_id)
P<-merge(P, bm_p, by="ensembl_transcript_id", all.x=TRUE)
nrow(P) #some double rows were introduced

P = P[match(Prom$name, P$name), ]

length(unique(P$ensembl_gene_id)) #number of unique genes

table(P[,c('strand.x','strand.y')]) #almost all strand listings are consistent

P<-P[, colnames(P)!='strand.y']
colnames(P)[colnames(P)=='strand.x'] = "strand"



## to be used by CGtools as the complete set of TSS's
peaks = data.frame(seqname=P$chr,
                   start=P$tss,
                   end=P$tss,
                   strand=P$strand)


Pseud<-min(P$SuRE[P$SuRE>0], na.rm=TRUE)/2
P$SuRE<-P$SuRE+Pseud
P$SuRE<-log10(P$SuRE)
PseudGro<-min(P$GROcap[P$GROcap>0], na.rm=TRUE)/2
P$GROcap<-P$GROcap+PseudGro
P$GROcap<-log10(P$GROcap)
PseudCage<-min(P$CAGE[P$CAGE>0], na.rm=TRUE)/2
P$CAGE<-P$CAGE+PseudCage
P$CAGE<-log10(P$CAGE)

#then calculate running mean for iLAD promoters:
P<-P[order(P$SuRE,sample(c(1:nrow(P)))),] #sort by SuRE and then random for ties
n<-60 #number of windows
w<-501 #window width (number of datapoints); if n*w > nrow(P) then windows overlap
s<-round(seq(from=w/2+0.0001, to=nrow(P)-w/2, length.out=n))
RM<-data.frame(SuRE.low=rep(NA,n), SuRE.mean=rep(NA,n), SuRE.hi=rep(NA,n), GROcap.lad=rep(NA,n), GROcap.ilad=rep(NA,n))
RM$SuRE.low=P$SuRE[s-floor(w/2)]
for(i in 1:n){RM$SuRE.mean[i]=mean(P$SuRE[(s[i]-floor(w/2)):(s[i]+floor(w/2))], na.rm=TRUE)}
RM$SuRE.hi=P$SuRE[s+floor(w/2)]
for(i in 1:n)
  {t<-P[(s[i]-floor(w/2)):(s[i]+floor(w/2)),]
   RM$GROcap.lad[i]<-mean(t$GROcap[t$LAD==1], na.rm=TRUE)
   RM$GROcap.ilad[i]<-mean(t$GROcap[t$LAD==0], na.rm=TRUE)
  }


#add first datapoint (SuRE equals pseudocount)
RM1<-RM[0,] #empty df
RM1[1,]<-c(rep(log10(Pseud),3), mean(P$GROcap[P$SuRE==log10(Pseud) & P$LAD==1]), mean(P$GROcap[P$SuRE==log10(Pseud) & P$LAD==0]))
RM<-rbind(RM1, RM)
rm(RM1)

#finally calculate LRS for all promoters:
P$LRS<- P$GROcap - approx(x=RM$SuRE.mean, y=RM$GROcap.ilad, xout=P$SuRE, rule=2)$y
#so the more negative the score, the more 'repressed' is the promoter by its chromatin/LAD context


#(arbitrary) cutoffs to define three groups of promoters:
INACT<- P$SuRE< -0.3 & P$LAD==1 & P$GROcap< -2 #inactive
NREP<- P$SuRE> 0 & P$LRS> -0.5 & P$LAD==1 & P$GROcap> -2 #not repressed
REP<- P$SuRE> 0.3 & P$LRS< -1 & P$LAD==1  & P$GROcap< -2 #repressed
Pcnts<-c(length(which(REP)), length(which(NREP)), length(which(INACT)))
names(Pcnts)<-c("repressed", "escaper", "inactive")
BND <- P$LAD==1 & !INACT & !NREP & !REP

#add class annotation column to P:
P$class<-NA
P$class[P$LAD==0]<-"iLAD"
P$class[INACT]<-"inactive"
P$class[NREP]<-"escaper"
P$class[REP]<-"repressed"
P$class[BND] <- "boundary"
P$class = factor(P$class, levels=c('iLAD', 'escaper', 'repressed', 'inactive', 'boundary'))

COLi<-"#00BBFF11" #dot color for iLAD promoters
COL_lad<-c("#FF0000", "#0077FF")
names(COL_lad)<-c('LAD', 'iLAD')

#color vector for plotting:
COL_class<-c("#A020F0", "#FFA500", "#006400", "#7e7e7e", "#0077FF")
names(COL_class)<-c("repressed", "escaper", "inactive", 'boundary', 'iLAD')

COL<-c("#A020F0", "#FFA500", "#006400")
names(COL)<-c("repressed", "escaper", "inactive")

class_names = paste0(levels(P$class), '; n=',table(P$class))
names(class_names) = levels(P$class)
P$class_n_prom = P$class
levels(P$class_n_prom) = class_names
COL_class_n_prom = COL_class[names(class_names)]
names(COL_class_n_prom) = class_names

lad_names = c(LAD=paste0('LAD; n=', table(P$LAD)['1']),
              iLAD=paste0('LAD; n=', table(P$LAD)['0']))
P$lad_n_prom = factor(ifelse(P$LAD==1, lad_names['LAD'], lad_names['iLAD']))
COL_lad_n = COL_lad
names(COL_lad_n) = lad_names


RM_melt = melt(RM, measure.vars=c('GROcap.ilad', 'GROcap.lad'))
RM_melt$variable = ifelse(RM_melt$variable=='GROcap.lad', lad_names['LAD'], lad_names['iLAD'])
ggplot(P, aes(x=SuRE, y=GROcap, color=lad_n_prom)) +
    geom_point(data=P[P$LAD==0, ], size=0.5, alpha=0.05) + 
    geom_point(data=P[P$LAD==1, ], size=0.5, alpha=0.2) + 
    theme_bw() +
    geom_line(data=RM_melt, aes(x=SuRE.mean, y=value, color=variable), size=1) +
    labs(y='log10(GROcap)', x='log10(SuRE)') +
    theme(legend.title=element_blank()) +
    scale_color_manual(values=COL_lad_n)
p_classes = P[which(P$class %in% c('inactive', 'escaper', 'repressed')),]
ggplot(P, aes(x=SuRE, y=GROcap)) + 
    geom_point(size=0.1,color=COLi) + 
    geom_point(data=p_classes, aes(color=class_n_prom), size=0.6) + 
    labs(y='log10(GROcap)', x='log10(SuRE)') +
    theme_bw() +
    geom_line(data=RM, aes(x=SuRE.mean, y=GROcap.ilad), color=COL_lad['iLAD']) +
    theme(legend.title=element_blank()) +
    scale_colour_manual(values=COL_class_n_prom) 

```

**conclusion:**

We now have a definition of 3 different groups. We can now look at what is different between these groups.

## create TF table

```{r}
bzip_table = read.csv('../raw_data/cl20170504_uniprot_human_bzip_ps50217.csv',
                      header=T, stringsAsFactors=F, row.names=1)
bzip_table$family = 'bZIP'
ets_table = read.csv('../raw_data/cl20170504_uniprot_human_ets_ps00346.csv',
                     header=T, stringsAsFactors=F, row.names=1)
ets_table$family = 'ETS'
tf_table = rbind(bzip_table, ets_table)
transcript_id = gsub('.*(ENST[0-9]+).*', '\\1', tf_table$Ensembl.transcript)
humanMart = useMart(biomart = 'ensembl', dataset = 'hsapiens_gene_ensembl')
bm_ens_gene = getBM(attributes=c('ensembl_gene_id', 'ensembl_transcript_id'),
                    filters='ensembl_transcript_id', values=transcript_id, mart=humanMart)

tf_table$ensembl_gene_id = bm_ens_gene$ensembl_gene_id[match(transcript_id,
                                                             bm_ens_gene$ensembl_transcript_id)]

```



```{r}
count_rep1 = read.table('../raw_data/expression/K562_rep1ReadsPerGene.out.tab')
count_rep2 = read.table('../raw_data/expression/K562_rep2ReadsPerGene.out.tab')

count_table = cbind(count_rep1[,4], rep2=count_rep2[,4])
rownames(count_table) = count_rep1[,1]
colnames(count_table) = c('rep1', 'rep2')
exp = SummarizedExperiment(assays = list(counts=count_table[-(1:4), ]))
dds = DESeqDataSet(exp, design= ~ 1)
fpm = fpm(dds)
fpm = rowMeans(fpm)
fpm = log10(fpm + min(fpm[fpm!=0])/2)

g_match = match(P$ensembl_gene_id, names(fpm))
P$K562_fpm = NaN
P$K562_fpm[!is.na(g_match)] = fpm[g_match[!is.na(g_match)]]
tf_table$K562_fpm = fpm[tf_table$ensembl_gene_id]
```

```{r}
wilcox_affinity <- function(x_affinity, y_affinity, groups, tf_table, id_vec=NULL){
  if (is.null(id_vec)){
    id_vec = colnames(x_affinity)
  }
  fit = mclapply(id_vec, function(id){
    x = x_affinity[,id]
    y = y_affinity[,id]
    r = rank(c(x,y))
    r_x = r[1:length(x)]
    r_y = r[-(1:length(x))]
    mean_r = c(mean(r_x), mean(r_y))
    direction = groups[which(mean_r==max(mean_r))]
    if (length(direction) == 2){
      direction = 'unchanged'
    }
    median_fc = median(x) / median(y)
    mean_fc = mean(x) / mean(y)
    rank_fc = mean_r[1] / mean_r[2]
    if (length(which(x==0))/length(x)>0.3 &
        length(which(y==0))/length(y)>0.3){
      w = chisq.test(rbind(table(x==0), table(y==0)))
    } else {
      w = wilcox.test(x, y)
    }
    return(list(w,direction, median_fc, mean_fc, rank_fc))
  })
  p_vec = unlist(lapply(fit, function(x){ x[[1]]$p.value}))
  p_adjust = p.adjust(p_vec, method='fdr')
  direction = lapply(fit, function(x){ x[[2]]})
  median_fc =  lapply(fit, function(x){ x[[3]]})
  mean_fc =  lapply(fit, function(x){ x[[4]]})
  rank_fc =  lapply(fit, function(x){ x[[5]]})
  result_table = cbind(id=id_vec, 
                       tf_table[id_vec, ],
                       direction=unlist(direction),
                       p_adjust = p_adjust,
                       median_fc = unlist(median_fc),
                       mean_fc = unlist(mean_fc), 
                       rank_fc = unlist(rank_fc), stringsAsFactors=F)
  return(result_table)
}

load('../workspace/cl20170505_transcript_ETS-bZIP_correlations_df.RData')

matched_evsr = matchSet(P[P$class%in%c('repressed', 'escaper'), ], 'class', 'escaper', 'SuRE')
escaper_cor_m = tf_cor_df[matched_evsr[which(matched_evsr$class=='escaper'),
                                       'ensembl_transcript_id'],]
repressed_cor_m = tf_cor_df[matched_evsr[which(matched_evsr$class=='repressed'),
                                       'ensembl_transcript_id'],]
evsr_cor_m = wilcox_affinity(escaper_cor_m, repressed_cor_m, c('escaper', 'repressed'), tf_table)
evsr_cor_sig = evsr_cor_m[which(evsr_cor_m$p_adjust < 0.5), ]
evsr_cor_sig[order(evsr_cor_sig$rank_fc),c('Gene.names...primary..', 'direction', 'p_adjust', 'rank_fc', 'K562_fpm')]

matched_evsilad = matchSet(P[P$class%in%c('iLAD', 'escaper'), ], 'class', 'escaper', 'SuRE')
escaper_cor_m = tf_cor_df[matched_evsilad[which(matched_evsilad$class=='escaper'),
                                       'ensembl_transcript_id'],]
ilad_cor_m = tf_cor_df[matched_evsilad[which(matched_evsilad$class=='iLAD'),
                                       'ensembl_transcript_id'],]
evsilad_cor_m = wilcox_affinity(escaper_cor_m, ilad_cor_m, c('escaper', 'ilad'), tf_table)
evsilad_cor_sig = evsilad_cor_m[which(evsilad_cor_m$p_adjust < 0.5), ]
evsilad_cor_sig[order(evsilad_cor_sig$rank_fc),c('Gene.names...primary..', 'direction', 'p_adjust', 'rank_fc', 'K562_fpm')]
```


## ATF2 knockdown
maybe ATF2 could be a good candidate, let's look at a knockdown experiment.

```{r}
atf2_kd = read.table('../raw_data/GSE50165_non-normalized.txt.gz', stringsAsFactors=F, header=T, sep='\t')
atf2_kd$fc = log2(rowMeans(atf2_kd[,c('ATF2_1.AVERAGE.SIGNAL', 'ATF2_2.AVERAGE_Signal')]) /
                  rowMeans(atf2_kd[,c('NT1_AVERAGE_Signal', 'NT2_AVERAGE_Signal')]))
escaper_fc_m = atf2_kd$fc[atf2_kd$SYMBOL %in% 
                          matched_evsilad[which(matched_evsilad$class=='escaper'),
                                          'name2']]
ilad_fc_m = atf2_kd$fc[atf2_kd$SYMBOL %in% 
                       matched_evsilad[which(matched_evsilad$class=='iLAD'),
                                       'name2']]
e_data = data.frame(class='escaper', fc = escaper_fc_m)
ilad_data = data.frame(class='iLAD', fc = ilad_fc_m)
ggplot(rbind(e_data, ilad_data), aes(x=class, y=fc)) +
    geom_violin(alpha=0.5) +
    geom_point(position=position_jitter(width=0.5), alpha=0.3, size=0.3) +
    scale_color_manual(values=COL_class)

## got a ETV4 KD and over expression dataset
## paper (https://dx.doi.org/10.1177%2F1947601910395578) published a list of genes higher than 2-fold 
## down and up regulated genes for knockdown and over expresion repectively in their supplementary data
etv4_kd_oe = read.table('../raw_data/etv4_kd_overexpression_highfc.csv', stringsAsFactors=F, header=T, sep='\t')
any(P[P$class=='escaper','name2']%in% etv4_kd_oe)
any(P[P$class=='escaper','name']%in% etv4_kd_oe)

```


```{r}
load('../workspace/cl20170426_transcript_correlations_df.RData')

matched_evsr = matchSet(P[P$class%in%c('repressed', 'escaper'), ], 'class', 'escaper', 'SuRE')
escaper_cor_m = cor_df[matched_evsr[which(matched_evsr$class=='escaper'),
                                       'ensembl_transcript_id'],]
repressed_cor_m = cor_df[matched_evsr[which(matched_evsr$class=='repressed'),
                                       'ensembl_transcript_id'],]
rownames(P) = P$ensembl_transcript_id
evsr_cor_m = wilcox_affinity(escaper_cor_m, repressed_cor_m, c('escaper', 'repressed'), P[,c('name2', 'CAGE', 'GROcap', 'K562_fpm')])
evsr_cor_sig = evsr_cor_m[which(evsr_cor_m$p_adjust < 0.5), ]
o = order(evsr_cor_sig$rank_fc)
evsr_cor_sig[o[c(0:20, (length(o)-20):length(o))],]

matched_evsilad = matchSet(P[P$class%in%c('iLAD', 'escaper'), ], 'class', 'escaper', 'SuRE')
escaper_cor_m = cor_df[matched_evsilad[which(matched_evsilad$class=='escaper'),
                                       'ensembl_transcript_id'],]
ilad_cor_m = cor_df[matched_evsilad[which(matched_evsilad$class=='iLAD'),
                                       'ensembl_transcript_id'],]
evsilad_cor_m = wilcox_affinity(escaper_cor_m, ilad_cor_m, c('escaper', 'ilad'),  P[,c('name2', 'CAGE', 'GROcap', 'K562_fpm')])
evsilad_cor_sig = evsilad_cor_m[which(evsilad_cor_m$p_adjust < 0.5), ]

o = order(evsilad_cor_sig$rank_fc)
evsilad_cor_sig[o[c(0:20, (length(o)-20):length(o))],]
```


```{r}
ets_bzip_interactions1 = string_data[string_data$node1%in%ets_table$Gene.names...primary.. & string_data$node2%in%bzip_table$Gene.names...primary.., ]
ets_bzip_interactions2 = string_data[string_data$node1%in%bzip_table$Gene.names...primary.. & string_data$node2%in%ets_table$Gene.names...primary.., ]
ets_bzip_interactions = rbind(ets_bzip_interactions1, ets_bzip_interactions2)

ets_bzip_list = unlist(ets_bzip_interactions[,c(1,2)])
table(ets_bzip_list[ets_bzip_list%in%bzip_table$Gene.names...primary..])
table(ets_bzip_list[ets_bzip_list%in%ets_table$Gene.names...primary..])
```


## PMA as drug to induce bZIPs

I found some data where they used PMA to recapitulate proliferation in K562 cells (GSE12736). Some bZIPs are said to be affected here. In addition some ETS-factors 

```{r}
test = read.csv('../raw_data/GSE12736_Limb-2-nov132008.csv', header=T, stringsAsFactors=F)
match_vec = match(P$name2, test$Symbol)
p_test = test[match_vec[!is.na(match_vec)], ]
p_test$class = P$class[!is.na(match_vec)]

for (cond in grep('Lfold', colnames(p_test),value=T)){
    print(ggplot(p_test[p_test$class!='boundary',], aes_string(x='class', y=cond, color='class')) +
            geom_violin(alpha=0.5) +
            geom_point(data=p_test[p_test$class%in%names(COL),], position=position_jitter(width=0.5), alpha=0.3, size=0.3) +
            scale_color_manual(values=COL_class))

}
match_vec = match(bzip_table$Gene.names...primary.. , test$Symbol)
bzip_test = test[match_vec[!is.na(match_vec)], ]
bzip_test$class = 'bZIP'

match_vec = match(ets_table$Gene.names...primary.. , test$Symbol)
ets_test = test[match_vec[!is.na(match_vec)], ]
ets_test$class = 'ETS'

p_ets_bzip_test = rbind(p_test, bzip_test, ets_test)
point_vec = c(names(COL), 'bZIP', 'ETS')
for (cond in grep('Lfold', colnames(p_test),value=T)){
    print(ggplot(p_ets_bzip_test[p_ets_bzip_test$class!='boundary',], aes_string(x='class', y=cond, color='class')) +
            geom_violin(alpha=0.5) +
            geom_point(data=p_ets_bzip_test[p_ets_bzip_test$class%in%point_vec,], position=position_jitter(width=0.5), alpha=0.3, size=0.3))
}
matched_evsilad = matchSet(P[P$class%in%c('iLAD', 'escaper'), ], 'class', 'escaper', 'GROcap')
match_vec = match(matched_evsilad$name2, test$Symbol)
matched_test = test[match_vec[!is.na(match_vec)], ]
matched_test$class = matched_evsilad$class[!is.na(match_vec)]
p_ets_bzip_test = rbind(matched_test, bzip_test, ets_test)
point_vec = c(names(COL), 'bZIP', 'ETS')
for (cond in grep('Lfold', colnames(p_test),value=T)){
    print(ggplot(p_ets_bzip_test, aes_string(x='class', y=cond, color='class')) +
            geom_violin(alpha=0.5) +
            geom_point(data=p_ets_bzip_test, position=position_jitter(width=0.5), alpha=0.3, size=0.3))
}

```

```{r}

escape_trip = c('ENST00000396668', 'ENST00000297347', 'ENST00000394226','ENST00000456315')

colnames(tf_cor_df) = tf_table[colnames(tf_cor_df), 'Gene.names...primary..']

this_cor = tf_cor_df[rownames(tf_cor_df)%in%escape_trip,]
for (id in escape_trip){
    o = order(this_cor[id,], decreasing=T)
    data = data.frame(value=unlist(this_cor[id,]),
               symbol=colnames(this_cor))
    print(head(t(this_cor[id,o]), n=20))
}

cor_rank = apply(this_cor,1, order)
rownames(cor_rank) = colnames(this_cor)
cor_order = rownames(cor_rank)[order(rowSums(cor_rank), decreasing=T)]
t(this_cor)[cor_order[1:20], ]



cor_rank = apply(tf_cor_df, 2, order)
rownames(cor_rank) = rownames(tf_cor_df)
colnames(cor_rank) = colnames(tf_cor_df)
this_rank = cor_rank[escape_trip, ]
order_vec = colnames(this_rank)[order(colSums(this_rank), decreasing=T)]
t(tf_cor_df[escape_trip, ])[order_vec[1:20], ]
t(tf_cor_df[escape_trip, ])[order(tf_cor_df['ENST00000394226', ]), ]

aff_table_jaspar = read.table('../raw_data/jaspar_affinity_300_300/seq_psam.dat', stringsAsFactors=F)
colnames(tf_table_jaspar) = c('name', 'species', 'class', 'family')
tf_table_jaspar = read.table('../raw_data/tf_table.txt', sep='\t', row.names=1, stringsAsFactors=F)
tf_translation = read.table('../raw_data/tf_translation.txt', stringsAsFactors=F)

tf_translation$K562_fpm = fpm[tf_translation$gene_id]
id_vec = colnames(aff_table_jaspar) = gsub('.xml','', colnames(aff_table_jaspar))

rank_table = apply(aff_table_jaspar,2,order)
rownames(rank_table) = rownames(aff_table_jaspar)
escape_name = P[escape_trip, 'name']
this_rank = rank_table[escape_name,]

order_vec = colnames(this_rank)[order(colSums(this_rank), decreasing=T)]

trip_aff = cbind.data.frame(symbol=tf_table_jaspar[order_vec, 'V2'],
                        t(aff_table_jaspar[escape_name, order_vec]) )

trip_aff[1:20,]
```