# knitr document van Steensel lab

# Gene repression in LADs
## Christ Leemans, 03-11-2016 - 26-07-2017

## Introduction
Generally speaking, genes inside lamina associated domains are not or very lowly expressed. These genes can either be actively repressed by their DNA context (e.g. heterochromatin, lamina association), or simply be inactive (because essential factors for expression are missing?). Yet another group of genes seem to evade gene repression in the context of lamina associated domains. In this report I would like to examine chromatin marks around TSS and TES of these different types of promoters.

## library and data loading
```{r, fig.width=10, fig.height=10, echo=FALSE, fig.width=10, fig.height=10}
library(reshape2)
library(rtracklayer)
library(ggplot2)
library(gridExtra)
library(plyr)
library(scales)
library(grid)
library(gtable)
library(affy)
library(limma)
library(biomaRt)
## FROM STACKOVERFLOW:
## https://stackoverflow.com/questions/12539348/ggplot-separate-legend-and-plot
g_legend<-function(a.gplot){
    tmp <- ggplot_gtable(ggplot_build(a.gplot))
    leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
    legend <- tmp$grobs[[leg]]
    return(legend)
}



## get a table with matching sets
## table = complete table to take matching sets from
## class_col = column name of class of interest
## class = name of class to match the set on
## order_on = column name to order on
matchSet <- function(table, class_col, class, order_on){
  o_vec = order(table[,order_on])
  o_table = table[o_vec, ]
  setA = which(o_table[,class_col]==class)
  setB = c(setA + 1, setA -1)
  ## check if setB is all within the possible indexes
  setB = setB[setB %in% 1:length(o_vec)]
  ## can also return o_table[unique(c(setA, setB)), ]
  ## but this way order is perserved.
  i_vec = o_vec[unique(c(setA, setB))]
  return(table[i_vec[order(i_vec)], ])
}


COLi<-"#00BBFF11" #dot color for iLAD promoters
COL_lad<-c("#FF0000", "#0077FF")
names(COL_lad)<-c('LAD', 'iLAD')

#color vector for plotting:
COL_class<-c("#A020F0", "#FFA500", "#006400", "#7e7e7e", "#0077FF")
names(COL_class)<-c("repressed", "escaper", "inactive", 'boundary', 'iLAD')

COL<-c("#A020F0", "#FFA500", "#006400")
names(COL)<-c("repressed", "escaper", "inactive")



id_table = read.table('../raw_data/transcript.table', stringsAsFactors=F,
                      row.names=1, col.names=c('transcript_id', 'gene_id',
                                               'symbol'))
load("~joris/mydata/git/SuRE/Joris//analysis_postNBT/Gencode_DF_generation_170707/gencode.sure.170712.rda")


P<-gencode.sure.170712[,c('chr', 'strand', 'txStart', 'txEnd', 'name', 'name2',
                          'tss', 'distance.to.previous.tss',
                          'k562.combined.45.55.sense',
                          'HT1080.sense', 'gro.cap.1kb.sense',
                          'encode.cage.reprocessed.1kb.sense',
                          'nr.of.tissues.in.which.expressed.max')]

names(P)[9:13]<-c("SuRE_K562", "SuRE_HT1080", "GROcap_K562", "CAGE_K562_encode", 'tissues_expressed')

rownames(P) = P$name

P$gene_id = id_table[P$name,'gene_id']


most_active <- function(P){
    result = ddply(P, .(gene_id), function(x){
        if (nrow(x)==1){
            result = x[1,]
        } else {
            result = x[order(x$SuRE_K562, decreasing=T)[1],]
        }
        return(result)
    })
    rownames(result) = result$name
    return(result)
}

p_most_active = most_active(P)


p_other = P[!rownames(P)%in%rownames(p_most_active), ]
p_new_names = rownames(p_most_active)
while (nrow(p_other) > 0){
    p_new = P[p_new_names, ]
    active_gr = makeGRangesFromDataFrame(data.frame(seqnames=p_new$chr,
                                                   start=p_new$tss,
                                                   end=p_new$tss,
                                                   strand=p_new$strand),
                                                   keep.extra.columns=TRUE)
    other_gr = makeGRangesFromDataFrame(data.frame(seqnames=p_other$chr,
                                                   start=p_other$tss,
                                                   end=p_other$tss,
                                                   strand=p_other$strand),
                                                   keep.extra.columns=TRUE)
    o = findOverlaps(active_gr,other_gr, maxgap=500, ignore.strand=FALSE)
    sub_o = o[p_new[queryHits(o), 'gene_id'] == p_other[subjectHits(o), 'gene_id']]
    p_other = p_other[-subjectHits(sub_o), ]
    p_active = most_active(p_other)
    p_other = p_other[!rownames(p_other)%in%rownames(p_active), ]
    p_new_names = c(p_new_names, rownames(p_active))
}

p_complete = rownames(P)

P = P[rownames(P)%in%p_new_names, ]



gene_gr <-makeGRangesFromDataFrame(data.frame(seqnames=P$chr,
                                              start=P$txStart,
                                              end=P$txEnd,
                                              strand=P$strand,
                                              name=P$name,
                                              tss=P$tss),
                                              keep.extra.columns=TRUE)
names(gene_gr) = P$name
tss_gr = gene_gr
ranges(tss_gr) = IRanges(gene_gr$tss,
                         gene_gr$tss)
names(tss_gr) = P$name

LAD_K562 = import.bed('~c.leemans/mydata/data/tracks/hg19/cl20161019_LAD_continuous_2state_K562.bed')
## to keep with Joris's previous analysis, let's assign a state to every promoter
o = findOverlaps(tss_gr, LAD_K562[LAD_K562$name=='LAD'])
P$LAD_K562 = 0
P$LAD_K562[queryHits(o)] = 1

pseudo_log10 <- function(val_vec){
    Pseud=min(val_vec[val_vec > 0], na.rm=TRUE)/2
    val_vec = val_vec + Pseud
    return(log10(val_vec))
}
for (col in c('SuRE_K562', 'SuRE_HT1080', 'GROcap_K562')){
    P[,col] = pseudo_log10(P[,col])
}


create_RM <-function(data, x, y, lad){
    #then calculate running mean for iLAD promoters:
    #sort by SuRE and then random for ties
    o = order(data[,x],sample(c(1:nrow(data))))

    x_sorted = data[o,x]
    y_sorted = data[o,y]
    lad_sorted = data[o,lad]

    n<-60 #number of windows
    w<-501 #window width (number of datapoints); if n*w > nrow(P) then windows overlap
    s<-round(seq(from=w/2+0.0001, to=nrow(data)-w/2, length.out=n))
    RM<-data.frame(x.low=rep(NA,n), x.mean=rep(NA,n), x.hi=rep(NA,n), y.lad=rep(NA,n), y.ilad=rep(NA,n))
    RM$x.low=x_sorted[s-floor(w/2)]
    for(i in 1:n){RM$x.mean[i]=mean(x_sorted[(s[i]-floor(w/2)):(s[i]+floor(w/2))], na.rm=TRUE)}
    RM$x.hi=x_sorted[s+floor(w/2)]
    for(i in 1:n)
      {t<-data.frame(LAD=lad_sorted[(s[i]-floor(w/2)):(s[i]+floor(w/2))],
                     y=y_sorted[(s[i]-floor(w/2)):(s[i]+floor(w/2))])
       RM$y.lad[i]<-mean(t$y[t$LAD==1], na.rm=TRUE)
       RM$y.ilad[i]<-mean(t$y[t$LAD==0], na.rm=TRUE)
      }
    #add first datapoint (SuRE equals pseudocount)
    RM1<-RM[0,] #empty df
    RM1[1,]<-c(rep(min(x_sorted),3), mean(y_sorted[x_sorted==min(x_sorted) & lad_sorted==1]), mean(y_sorted[x_sorted==min(x_sorted) & lad_sorted==0]))
    RM<-rbind(RM1, RM)
    rm(RM1)
    return(RM)
}

RM_GRO = create_RM(P, 'SuRE_K562', 'GROcap_K562', lad='LAD_K562')

P$LRS_GROcap<- P$GROcap_K562 - approx(x=RM_GRO$x.mean, y=RM_GRO$y.ilad, xout=P$SuRE_K562, rule=2)$y


classify <- function(sure, exp, lrs, lad, exp_cut){
    INACT<- sure< -0.3 & lad & exp< exp_cut #inactive
    NREP<- sure> 0 & lrs > -0.5 & lad & exp> exp_cut #not repressed
    REP<- sure> 0.3 & lrs < -1 & lad  & exp< exp_cut #repressed
    Pcnts<-c(length(which(REP)), length(which(NREP)), length(which(INACT)))
    names(Pcnts)<-c("repressed", "escaper", "inactive")
    BND <- lad & !INACT & !NREP & !REP
    class = rep(NA, length(sure))
    class[lad==0] = 'iLAD'
    class[INACT]<-"inactive"
    class[NREP]<-"escaper"
    class[REP]<-"repressed"
    class[BND] <- "boundary"
    return(factor(class, levels=c('iLAD', 'escaper', 'repressed', 'inactive', 'boundary')))
}

P$class_GROcap = classify(P$SuRE_K562, P$GROcap_K562, P$LRS_GROcap, P$LAD_K562, -2)

```


```
for chip in $(ls ~/mydata/data/ChIP/hist/*)
do
    chip_name=$(basename "${chip%.*}")
    out_tss=raw_data/$chip_name"_tss.txt.gz"
    out_tes=raw_data/$chip_name"_tes.txt.gz"
    nice -19 computeMatrix reference-point -R raw_data/transcripts_hg19.bed \
                                           -S $chip  \
                                           --referencePoint TSS \
                                           -a 22000 -b 22000 \
                                           --binSize=200 -p 20 \
                                           --averageTypeBins sum \
                                           --missingDataAsZero \
                                           --outFileName $out_tss
    nice -19 computeMatrix reference-point -R raw_data/transcripts_hg19.bed \
                                           -S $chip  \
                                           --referencePoint TES \
                                           -a 22000 -b 22000 \
                                           --binSize=200 -p 20 \
                                           --averageTypeBins sum \
                                           --missingDataAsZero \
                                           --outFileName $out_tes
    out_tssr=raw_data/$chip_name"_tssr.txt.gz"
    bwtool summary -header \
                   -fill=0 \
                   -with-sum \
                   -skip-median \
                   -keep-bed \
                   <(awk -vOFS='\t' '{
                       if ($6=="+"){
                           start = $2 - 50
                           end = $3 + 300
                       } else {
                           start = $2-300
                           end = $3 + 50
                       }
                       start = start > 0 ? start : 0 ;
                       end = end > 0 ? end : 0 ;
                       print $1, start, end, $4, 0, $6}' raw_data/tss_hg19.bed) \
                   $chip /dev/stdout | gzip -c > $out_tssr &
    out_body=raw_data/$chip_name"_body.txt.gz"
    bwtool summary -header \
                   -fill=0 \
                   -with-sum \
                   -skip-median \
                   -keep-bed \
                   <(awk -vOFS='\t' '{
                       if ($6=="+"){
                           start = $2 + 5000
                           end = $3 + 20000
                       } else {
                           start = $2 - 20000
                           end = $3 - 5000
                       }
                       start = start > 0 ? start : 0 ;
                       end = end > 0 ? end : 0 ;
                       print $1, start, end, $4, 0, $6}' raw_data/tss_hg19.bed) \
                   $chip /dev/stdout | gzip -c > $out_body
done
```


```{r, fig.width=10, fig.height=15}

group_mean <- function(name, import_list, P){
    data = import_list[[name]][,6:ncol(import_list[[name]])]
    data = data.frame(class=P$class_GROcap, data[P$name, ])
    mean_data = ddply(data, .(class), function(x){
            colMeans(x[,2:ncol(x)])
    })
    rownames(mean_data) = mean_data[,1]
    data.frame(t(mean_data[,-1]) / max(mean_data[,-1]), pos=-110:109*200, name=name)
}
group_center <- function(name, import_list, P){
  data = import_list[[name]][,6:ncol(import_list[[name]])]
  data = data.frame(class=P$class_GROcap,
                    value=rowSums(data[P$name, 109:113]),
                    name=name)
  data$value = data$value / max(data$value)
  return(data)
}



chip_vec = list.files('~/mydata/data/ChIP/hist/')
tss_file_vec = gsub('.bigWig', '_tss.txt.gz', chip_vec)
tes_file_vec = gsub('.bigWig', '_tes.txt.gz', chip_vec)
name_vec = gsub('K562_(.*)_hg19.*', '\\1', chip_vec)

tss_import_list = lapply(paste0('../raw_data/', tss_file_vec), read.table,
                         sep='\t', skip=1, stringsAsFactors=T, row.names=4)
tes_import_list = lapply(paste0('../raw_data/', tes_file_vec), read.table,
                         sep='\t', skip=1, stringsAsFactors=T, row.names=4)
names(tss_import_list) = names(tes_import_list) = name_vec

p_matched = matchSet(P[P$class%in%c('iLAD', 'escaper'), ], 'class_GROcap',
                     'escaper', 'GROcap_K562')
p_class = P[P$class_GROcap%in%c('escaper', 'repressed', 'inactive'),]

p_subset = rbind.data.frame(p_class, p_matched[p_matched$class_GROcap=='iLAD', ])
mean_list = lapply(name_vec, group_mean, import_list=tss_import_list,
                   P=p_subset)


class_data = do.call(rbind, mean_list)

ggplot(melt(class_data, id.vars=c('pos', 'name')),
       aes(x=pos, y=value, color=variable)) +
    geom_line() +
    ggtitle('histone modifications around transcription start site') +
    scale_color_manual(values=COL_class) +
    facet_wrap(~name, nrow=2)




mean_list = lapply(name_vec, group_mean, import_list=tes_import_list,
                   P=p_subset)


class_data = do.call(rbind, mean_list)

ggplot(melt(class_data, id.vars=c('pos', 'name')),
       aes(x=pos, y=value, color=variable)) +
    geom_line() +
    ggtitle('histone modifications around transcription termination site') +
    scale_color_manual(values=COL_class) +
    facet_wrap(~name)




```


```{r, fig.width=10, fig.height=30}

tssr_file_vec = gsub('.bigWig', '_tssr.txt.gz', chip_vec)
body_file_vec = gsub('.bigWig', '_body.txt.gz', chip_vec)

tssr_import_list = lapply(paste0('../raw_data/', tssr_file_vec), read.table,
                          row.names=4, stringsAsFactors=F,
                          col.names=c('chrom', 'start', 'end', 'name', 'score',
                                     'strand', 'size', 'num_data', 'min', 'max',
                                     'mean', 'sum'))
body_import_list = lapply(paste0('../raw_data/', body_file_vec), read.table,
                          row.names=4, stringsAsFactors=F,
                          col.names=c('chrom', 'start', 'end', 'name', 'score',
                                      'strand', 'size', 'num_data', 'min', 'max',
                                      'mean', 'sum'))

tssr_list = lapply(tssr_import_list,
                   function(tssr, P){
                       data.frame(mean=tssr[P$name, 'mean'], class=P$class_GROcap,
                                  region='tss region (-50bp:+300bp)',
                                  log10=pseudo_log10(tssr[P$name, 'mean']))
                   }, P=p_subset)

body_list = lapply(body_import_list,
                   function(body, P){
                       data.frame(mean=body[P$name, 'mean'], class=P$class_GROcap,
                                  region='gene body (+3000bp:+20000bp)',
                                  log10=pseudo_log10(body[P$name, 'mean']))
                   }, P=p_subset)

names(tssr_list) = names(body_list) = name_vec

data_list = lapply(name_vec, function(name){
                       data = rbind.data.frame(tssr_list[[name]],
                                               body_list[[name]])
                       data$chip_name = name
                       return(data)
                   })

data = do.call(rbind.data.frame, data_list)
p_list = list()
for (name in name_vec){
    p_list[[name]] = ggplot(data[data$chip_name==name, ],
                            aes(y=log10, x=class, color=class)) +
                        geom_violin() +
                        ggtitle(name) +
                        geom_point(alpha=0.4, size=0.4,
                                   position=position_jitter(width=0.1)) +
                        facet_wrap(~region) +
                        scale_color_manual(values=COL_class)
}
do.call(grid.arrange, c(p_list, ncol=2))
```
