# knitr document van Steensel lab

# Gene repression in LADs
## Christ Leemans, 03-11-2016 - 26-07-2017

## Introduction
Generally speaking, genes inside lamina associated domains are not or very lowly expressed. These genes can either be actively repressed by their DNA context (e.g. heterochromatin, lamina association), or simply be inactive (because essential factors for expression are missing?). Yet another group of genes seem to evade gene repression in the context of lamina associated domains. In this report I would like to examine chromatin marks around TSS and TES of these different types of promoters.

## library and data loading
```{r, fig.width=10, fig.height=10, echo=FALSE, fig.width=10, fig.height=10}
library(reshape2)
library(rtracklayer)
library(ggplot2)
library(gridExtra)
library(plyr)
library(scales)
library(grid)
library(gtable)
library(affy)
library(limma)
library(biomaRt)
## FROM STACKOVERFLOW:
## https://stackoverflow.com/questions/12539348/ggplot-separate-legend-and-plot
g_legend<-function(a.gplot){
    tmp <- ggplot_gtable(ggplot_build(a.gplot))
    leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
    legend <- tmp$grobs[[leg]]
    return(legend)
}



## get a table with matching sets
## table = complete table to take matching sets from
## class_col = column name of class of interest
## class = name of class to match the set on
## order_on = column name to order on
matchSet <- function(table, class_col, class, order_on){
  o_vec = order(table[,order_on])
  o_table = table[o_vec, ]
  setA = which(o_table[,class_col]==class)
  setB = c(setA + 1, setA -1)
  ## check if setB is all within the possible indexes
  setB = setB[setB %in% 1:length(o_vec)]
  ## can also return o_table[unique(c(setA, setB)), ]
  ## but this way order is perserved.
  i_vec = o_vec[unique(c(setA, setB))]
  return(table[i_vec[order(i_vec)], ])
}


COLi<-"#00BBFF11" #dot color for iLAD promoters
COL_lad<-c("#FF0000", "#0077FF")
names(COL_lad)<-c('LAD', 'iLAD')

#color vector for plotting:
COL_class<-c("#A020F0", "#FFA500", "#006400", "#7e7e7e", "#0077FF")
names(COL_class)<-c("repressed", "escaper", "inactive", 'boundary', 'iLAD')

COL<-c("#A020F0", "#FFA500", "#006400")
names(COL)<-c("repressed", "escaper", "inactive")



id_table = read.table('../raw_data/transcript.table', stringsAsFactors=F,
                      row.names=1, col.names=c('transcript_id', 'gene_id',
                                               'symbol'))
load("~joris/mydata/git/SuRE/Joris//analysis_postNBT/Gencode_DF_generation_170707/gencode.sure.170712.rda")


P<-gencode.sure.170712[,c('chr', 'strand', 'txStart', 'txEnd', 'name', 'name2',
                          'tss', 'distance.to.previous.tss',
                          'k562.combined.45.55.sense',
                          'HT1080.sense', 'gro.cap.1kb.sense',
                          'encode.cage.reprocessed.1kb.sense',
                          'nr.of.tissues.in.which.expressed.max')]

names(P)[9:13]<-c("SuRE_K562", "SuRE_HT1080", "GROcap_K562", "CAGE_K562_encode", 'tissues_expressed')

rownames(P) = P$name

P$gene_id = id_table[P$name,'gene_id']


most_active <- function(P){
    result = ddply(P, .(gene_id), function(x){
        if (nrow(x)==1){
            result = x[1,]
        } else {
            result = x[order(x$SuRE_K562, decreasing=T)[1],]
        }
        return(result)
    })
    rownames(result) = result$name
    return(result)
}

p_most_active = most_active(P)


p_other = P[!rownames(P)%in%rownames(p_most_active), ]
p_new_names = rownames(p_most_active)
while (nrow(p_other) > 0){
    p_new = P[p_new_names, ]
    active_gr = makeGRangesFromDataFrame(data.frame(seqnames=p_new$chr,
                                                   start=p_new$tss,
                                                   end=p_new$tss,
                                                   strand=p_new$strand),
                                                   keep.extra.columns=TRUE)
    other_gr = makeGRangesFromDataFrame(data.frame(seqnames=p_other$chr,
                                                   start=p_other$tss,
                                                   end=p_other$tss,
                                                   strand=p_other$strand),
                                                   keep.extra.columns=TRUE)
    o = findOverlaps(active_gr,other_gr, maxgap=500, ignore.strand=FALSE)
    sub_o = o[p_new[queryHits(o), 'gene_id'] == p_other[subjectHits(o), 'gene_id']]
    p_other = p_other[-subjectHits(sub_o), ]
    p_active = most_active(p_other)
    p_other = p_other[!rownames(p_other)%in%rownames(p_active), ]
    p_new_names = c(p_new_names, rownames(p_active))
}

p_complete = rownames(P)

P = P[rownames(P)%in%p_new_names, ]



gene_gr <-makeGRangesFromDataFrame(data.frame(seqnames=P$chr,
                                              start=P$txStart,
                                              end=P$txEnd,
                                              strand=P$strand,
                                              name=P$name,
                                              tss=P$tss),
                                              keep.extra.columns=TRUE)
names(gene_gr) = P$name
tss_gr = gene_gr
ranges(tss_gr) = IRanges(gene_gr$tss,
                         gene_gr$tss)
names(tss_gr) = P$name

LAD_K562 = import.bed('~c.leemans/mydata/data/tracks/hg19/cl20161019_LAD_continuous_2state_K562.bed')
## to keep with Joris's previous analysis, let's assign a state to every promoter
o = findOverlaps(tss_gr, LAD_K562[LAD_K562$name=='LAD'])
P$LAD_K562 = 0
P$LAD_K562[queryHits(o)] = 1

pseudo_log10 <- function(val_vec){
    Pseud=min(val_vec[val_vec > 0], na.rm=TRUE)/2
    val_vec = val_vec + Pseud
    return(log10(val_vec))
}
for (col in c('SuRE_K562', 'SuRE_HT1080', 'GROcap_K562')){
    P[,col] = pseudo_log10(P[,col])
}


create_RM <-function(data, x, y, lad){
    #then calculate running mean for iLAD promoters:
    #sort by SuRE and then random for ties
    o = order(data[,x],sample(c(1:nrow(data))))

    x_sorted = data[o,x]
    y_sorted = data[o,y]
    lad_sorted = data[o,lad]

    n<-60 #number of windows
    w<-501 #window width (number of datapoints); if n*w > nrow(P) then windows overlap
    s<-round(seq(from=w/2+0.0001, to=nrow(data)-w/2, length.out=n))
    RM<-data.frame(x.low=rep(NA,n), x.mean=rep(NA,n), x.hi=rep(NA,n), y.lad=rep(NA,n), y.ilad=rep(NA,n))
    RM$x.low=x_sorted[s-floor(w/2)]
    for(i in 1:n){RM$x.mean[i]=mean(x_sorted[(s[i]-floor(w/2)):(s[i]+floor(w/2))], na.rm=TRUE)}
    RM$x.hi=x_sorted[s+floor(w/2)]
    for(i in 1:n)
      {t<-data.frame(LAD=lad_sorted[(s[i]-floor(w/2)):(s[i]+floor(w/2))],
                     y=y_sorted[(s[i]-floor(w/2)):(s[i]+floor(w/2))])
       RM$y.lad[i]<-mean(t$y[t$LAD==1], na.rm=TRUE)
       RM$y.ilad[i]<-mean(t$y[t$LAD==0], na.rm=TRUE)
      }
    #add first datapoint (SuRE equals pseudocount)
    RM1<-RM[0,] #empty df
    RM1[1,]<-c(rep(min(x_sorted),3), mean(y_sorted[x_sorted==min(x_sorted) & lad_sorted==1]), mean(y_sorted[x_sorted==min(x_sorted) & lad_sorted==0]))
    RM<-rbind(RM1, RM)
    rm(RM1)
    return(RM)
}

RM_GRO = create_RM(P, 'SuRE_K562', 'GROcap_K562', lad='LAD_K562')

P$LRS_GROcap<- P$GROcap_K562 - approx(x=RM_GRO$x.mean, y=RM_GRO$y.ilad, xout=P$SuRE_K562, rule=2)$y


classify <- function(sure, exp, lrs, lad, exp_cut){
    INACT<- sure< -0.3 & lad & exp< exp_cut #inactive
    NREP<- sure> 0 & lrs > -0.5 & lad & exp> exp_cut #not repressed
    REP<- sure> 0.3 & lrs < -1 & lad  & exp< exp_cut #repressed
    Pcnts<-c(length(which(REP)), length(which(NREP)), length(which(INACT)))
    names(Pcnts)<-c("repressed", "escaper", "inactive")
    BND <- lad & !INACT & !NREP & !REP
    class = rep(NA, length(sure))
    class[lad==0] = 'iLAD'
    class[INACT]<-"inactive"
    class[NREP]<-"escaper"
    class[REP]<-"repressed"
    class[BND] <- "boundary"
    return(factor(class, levels=c('iLAD', 'escaper', 'repressed', 'inactive', 'boundary')))
}

P$class_GROcap = classify(P$SuRE_K562, P$GROcap_K562, P$LRS_GROcap, P$LAD_K562, -2)

```


```

for chip in $(ls ~/mydata/data/ChIP/hist/*.bigWig)
do
    chip_name=$(basename "${chip%.*}")
    out_tss=raw_data/$chip_name"_tss.txt.gz"
    out_tes=raw_data/$chip_name"_tes.txt.gz"
    nice -19 computeMatrix reference-point -R raw_data/transcripts_hg19.bed \
                                           -S $chip  \
                                           --referencePoint TSS \
                                           -a 22000 -b 22000 \
                                           --binSize=200 -p 20 \
                                           --averageTypeBins mean \
                                           --missingDataAsZero \
                                           --outFileName $out_tss
    nice -19 computeMatrix reference-point -R raw_data/transcripts_hg19.bed \
                                           -S $chip  \
                                           --referencePoint TES \
                                           -a 22000 -b 22000 \
                                           --binSize=200 -p 20 \
                                           --averageTypeBins mean \
                                           --missingDataAsZero \
                                           --outFileName $out_tes
    out_tssr=raw_data/$chip_name"_tssr.txt.gz"
    bwtool summary -header \
                   -fill=0 \
                   -with-sum \
                   -skip-median \
                   -keep-bed \
                   <(awk -vOFS='\t' '{
                       if ($6=="+"){
                           start = $2 - 50
                           end = $3 + 300
                       } else {
                           start = $2-300
                           end = $3 + 50
                       }
                       start = start > 0 ? start : 0 ;
                       end = end > 0 ? end : 0 ;
                       print $1, start, end, $4, 0, $6}' raw_data/tss_hg19.bed) \
                   $chip /dev/stdout | gzip -c > $out_tssr &
    out_body=raw_data/$chip_name"_body.txt.gz"
    bwtool summary -header \
                   -fill=0 \
                   -with-sum \
                   -skip-median \
                   -keep-bed \
                   <(awk -vOFS='\t' '{
                       if ($6=="+"){
                           start = $2 + 5000
                           end = $3 + 20000
                       } else {
                           start = $2 - 20000
                           end = $3 - 5000
                       }
                       start = start > 0 ? start : 0 ;
                       end = end > 0 ? end : 0 ;
                       print $1, start, end, $4, 0, $6}' raw_data/tss_hg19.bed) \
                   $chip /dev/stdout | gzip -c > $out_body
    out_peaks=raw_data/$chip_name"_mean_peak.txt"
    peaks=$(echo $chip | sed 's/ENCF.*/replicated_peaks\*/')
    bwtool summary -header \
                   -with-sum \
                   -skip-median \
                   -total \
                   <(gunzip -c $peaks  | awk -vOFS='\t' '{print $1, $2, $3}') \
                   $chip /dev/stdout > $out_peaks
done



for chip in $(ls ~/mydata/data/ChIP/elongation_factor/*.bigWig)
do
    chip_name=$(basename "${chip%.*}")
    out_tss=raw_data/$chip_name"_tss.txt.gz"
    out_tes=raw_data/$chip_name"_tes.txt.gz"
    nice -19 computeMatrix reference-point -R raw_data/transcripts_hg19.bed \
                                           -S $chip  \
                                           --referencePoint TSS \
                                           -a 22000 -b 22000 \
                                           --binSize=200 -p 20 \
                                           --averageTypeBins mean \
                                           --missingDataAsZero \
                                           --outFileName $out_tss
    nice -19 computeMatrix reference-point -R raw_data/transcripts_hg19.bed \
                                           -S $chip  \
                                           --referencePoint TES \
                                           -a 22000 -b 22000 \
                                           --binSize=200 -p 20 \
                                           --averageTypeBins mean \
                                           --missingDataAsZero \
                                           --outFileName $out_tes
    out_tssr=raw_data/$chip_name"_tssr.txt.gz"
    bwtool summary -header \
                   -fill=0 \
                   -with-sum \
                   -skip-median \
                   -keep-bed \
                   <(awk -vOFS='\t' '{
                       if ($6=="+"){
                           start = $2 - 50
                           end = $3 + 300
                       } else {
                           start = $2-300
                           end = $3 + 50
                       }
                       start = start > 0 ? start : 0 ;
                       end = end > 0 ? end : 0 ;
                       print $1, start, end, $4, 0, $6}' raw_data/tss_hg19.bed) \
                   $chip /dev/stdout | gzip -c > $out_tssr &
    out_body=raw_data/$chip_name"_body.txt.gz"
    bwtool summary -header \
                   -fill=0 \
                   -with-sum \
                   -skip-median \
                   -keep-bed \
                   <(awk -vOFS='\t' '{
                       if ($6=="+"){
                           start = $2 + 5000
                           end = $3 + 20000
                       } else {
                           start = $2 - 20000
                           end = $3 - 5000
                       }
                       start = start > 0 ? start : 0 ;
                       end = end > 0 ? end : 0 ;
                       print $1, start, end, $4, 0, $6}' raw_data/tss_hg19.bed) \
                   $chip /dev/stdout | gzip -c > $out_body
done


for chip in $(ls ~/mydata/data/ChIP/elongation_factor/*.bigWig)
do
    chip_name=$(basename "${chip%.*}")
    out_tss=raw_data/$chip_name"_tss_11000.txt.gz"
    out_tes=raw_data/$chip_name"_tes_11000.txt.gz"
    nice -19 computeMatrix reference-point -R raw_data/transcripts_hg19.bed \
                                           -S $chip  \
                                           --referencePoint TSS \
                                           -a 5500 -b 5500 \
                                           --binSize=20 -p 20 \
                                           --averageTypeBins mean \
                                           --missingDataAsZero \
                                           --outFileName $out_tss
    nice -19 computeMatrix reference-point -R raw_data/transcripts_hg19.bed \
                                           -S $chip  \
                                           --referencePoint TES \
                                           -a 5500 -b 5500 \
                                           --binSize=20 -p 20 \
                                           --averageTypeBins mean \
                                           --missingDataAsZero \
                                           --outFileName $out_tes
done





```

```{r}

## to mask potential confounding activity of nearby genes from plots, I
## look at the nearest genes both sides and calculate the distance of
## the gene start site and end towards the midpoint between the neighbors

chrom_sizes = read.table('~/mydata/data/hg19/hg19.chrom.sizes', row.names=1)

gencode_gr = import.gff('~/mydata/data/tracks/hg19/gencode.v19.annotation.gff3.gz')
gencode_gr = gencode_gr[gencode_gr$type=='gene']

follow_vec = follow(gencode_gr, ignore.strand=T)
precede_vec = precede(gencode_gr, ignore.strand=T)

regions = data.frame(row.names = gencode_gr$ID,
                     strand = strand(gencode_gr))
left = ifelse(is.na(follow_vec), -Inf,
              (start(gencode_gr) - end(gencode_gr)[follow_vec])/2)
right = ifelse(is.na(precede_vec), Inf,
               (start(gencode_gr)[precede_vec] - end(gencode_gr))/2)

regions$start = ifelse(regions$strand=='+', -left, -right)
regions$end = ifelse(regions$strand=='+', right, left)

```



```{r, fig.width=10, fig.height=15}
mask_regions = read.table('../raw_data/transcript_region_masks.txt', header=T,
                     row.names=1, stringsAsFactors=F)

mask <- function(signal_data, mask_regions, start=-22000, end=22000, step=200,
                 is_tss=T){
    pos_vec = (start / step) : (end / step - 1) + .5
    data_list = mclapply(rownames(signal_data), function(t_id){
             if (is_tss){
                 mask = which(pos_vec < floor(mask_regions[t_id, 'start'] / step))
             } else {
                 mask = which(pos_vec > floor(mask_regions[t_id, 'end'] / step))
             }
             row = signal_data[t_id,]
             row[mask] = NA
             return(row)
         }, mc.cores=10)
    data = do.call(rbind.data.frame, data_list)
    colnames(data) = pos_vec * step
    return(data)
}

group_mean <- function(signal_data, P, class_name='class_GROcap',
                       start=-22000, end=22000, step=200){
    pos_vec = (start / step) : (end / step - 1) + .5
    data = data.frame(class=P[, class_name], signal_data[P$name, ])
    mean_data = ddply(data, .(class), function(x){
         apply(x[,2:ncol(x)],2,function(y){
             t = t.test(y)
             as.vector(c(t$estimate, t$conf.int))
         })
    })
    class_vec = unique(mean_data$class)
    i_vec = 1:length(class_vec) * 3
    result = data.frame(class=rep(class_vec, ncol(signal_data)),
                        mean=do.call(c, mean_data[i_vec-2, -1]),
                        min=do.call(c, mean_data[i_vec-1, -1]),
                        max=do.call(c, mean_data[i_vec, -1]),
                        pos=rep(pos_vec * step, each=length(class_vec)))
    return(result)
}

group_center <- function(name, import_list, P){
  data = import_list[[name]][,6:ncol(import_list[[name]])]
  data = data.frame(class=P$class_GROcap,
                    value=rowSums(data[P$name, 109:113]),
                    name=name)
  data$value = data$value / max(data$value)
  return(data)
}



chip_vec = list.files('~/mydata/data/ChIP/hist/', pattern='*.bigWig')
tss_file_vec = gsub('.bigWig', '_tss.txt.gz', chip_vec)
tes_file_vec = gsub('.bigWig', '_tes.txt.gz', chip_vec)
name_vec = gsub('K562_(.*)_hg19.*', '\\1', chip_vec)


tss_import_list = lapply(paste0('../raw_data/', tss_file_vec), read.table,
                         sep='\t', skip=1, stringsAsFactors=T, row.names=4)
tes_import_list = lapply(paste0('../raw_data/', tes_file_vec), read.table,
                         sep='\t', skip=1, stringsAsFactors=T, row.names=4)
names(tss_import_list) = names(tes_import_list) = name_vec

p_matched = matchSet(P[P$class_GROcap%in%c('iLAD', 'escaper'), ], 'class_GROcap',
                     'escaper', 'GROcap_K562')
p_class = P[P$class_GROcap%in%c('escaper', 'repressed', 'inactive'),]

p_subset = rbind.data.frame(p_class, p_matched[p_matched$class_GROcap=='iLAD', ])

```

## chip-seq signal around TSS

I took the regions around the TSS in 200bp bins and I calculated the mean fold change over control signal. This, I then normalized by the average signal on peaks shared between
the two replicates. Plots depict the mean over all promoters within each of the
LAD classes. The iLAD class is a matching set based on GROcap signal of the escaper
promoters.


```{r, fig.width=10, fig.height=15}

run_mask <- function(data, P, class_name='class_GROcap',
                       start=-22000, end=22000, step=200){
    mask(data[P$name, 6:ncol(data)], P,
         start=start, end=end, step=step)
}


run_group_mean <- function(name, mask_list, P, class_name='class_GROcap',
                       start=-22000, end=22000, step=200){
    result = data.frame(group_mean(mask_list[[name]], P, class_name=class_name,
                                   start=start, end=end, step=step),
                        name=name)
    result$mean = result$mean / max(result$mean)
    result$min = result$min / max(result$min)
    result$max = result$max / max(result$max)
    return(result)
}


tss_mask_list = lapply(tss_import_list, run_mask, p_subset)
tss_mean_list = lapply(name_vec, run_group_mean, tss_mask_list, p_subset)


class_data = do.call(rbind.data.frame, tss_mean_list)


ggplot(class_data, aes(x=pos, y=mean, color=class)) +
    geom_line() +
    ggtitle('histone modifications around transcription start site') +
    scale_color_manual(values=COL_class) +
    facet_wrap(~name, nrow=4)

```



## chip-seq signal around TES

I took the regions around the transcription end site in 200bp bins and I calculated
the mean fold change over control signal. This, I then normalized by the average signal
on peaks shared between the two replicates. Plots depict the mean over all promoters within
each of the LAD classes. The iLAD class is a matching set based on GROcap signal of the
escaper promoters.


```{r, fig.width=10, fig.height=15}

tes_mask_list = lapply(tes_import_list, run_mask, p_subset)
tes_mean_list = lapply(name_vec, run_group_mean, tes_mask_list, p_subset)


class_data = do.call(rbind.data.frame, tes_mean_list)
ggplot(melt(class_data, id.vars=c('pos', 'name')),
       aes(x=pos, y=value, color=variable)) +
    geom_line() +
    ggtitle('histone modifications around transcription termination site') +
    scale_color_manual(values=COL_class) +
    facet_wrap(~name, nrow=4)


```


```{r, fig.width=10, fig.height=30}

tssr_file_vec = gsub('.bigWig', '_tssr.txt.gz', chip_vec)
body_file_vec = gsub('.bigWig', '_body.txt.gz', chip_vec)
tssr_import_list = lapply(paste0('../raw_data/', tssr_file_vec), read.table,
                          row.names=4, stringsAsFactors=F,
                          col.names=c('chrom', 'start', 'end', 'name', 'score',
                                     'strand', 'size', 'num_data', 'min', 'max',
                                     'mean', 'sum'))
body_import_list = lapply(paste0('../raw_data/', body_file_vec), read.table,
                          row.names=4, stringsAsFactors=F,
                          col.names=c('chrom', 'start', 'end', 'name', 'score',
                                      'strand', 'size', 'num_data', 'min', 'max',
                                      'mean', 'sum'))

names(tssr_import_list) = names(body_import_list) = name_vec

for (name in name_vec){
    tssr_import_list[[name]]$mean = tssr_import_list[[name]]$mean /
                                    peak_import_list[[name]]$mean
    body_import_list[[name]]$mean = body_import_list[[name]]$mean /
                                    peak_import_list[[name]]$mean
}

tssr_list = lapply(tssr_import_list,
                   function(tssr, P){
                       data.frame(mean=tssr[P$name, 'mean'], class=P$class_GROcap,
                                  region='tss region (-50bp:+300bp)',
                                  log10=pseudo_log10(tssr[P$name, 'mean']))
                   }, P=p_subset)

body_list = lapply(body_import_list,
                   function(body, P){
                       data.frame(mean=body[P$name, 'mean'], class=P$class_GROcap,
                                  region='gene body (+3000bp:+20000bp)',
                                  log10=pseudo_log10(body[P$name, 'mean']))
                   }, P=p_subset)

names(tssr_list) = names(body_list) = name_vec

data_list = lapply(name_vec, function(name){
                       data = rbind.data.frame(tssr_list[[name]],
                                               body_list[[name]])
                       data$chip_name = name
                       return(data)
                   })

data = do.call(rbind.data.frame, data_list)

ggplot(data, aes(y=log10, x=class, color=class)) +
    geom_violin() +
    ggtitle(name) +
    geom_point(alpha=0.4, size=0.4,
               position=position_jitter(width=0.1)) +
    facet_wrap(~chip_name + region, ncol=4) +
    scale_color_manual(values=COL_class)
```


```{r}

chip_vec = list.files('~/mydata/data/ChIP/elongation_factor/', pattern='*.bigWig')
tss_file_vec = gsub('.bigWig', '_tss_11000.txt.gz', chip_vec)
tes_file_vec = gsub('.bigWig', '_tes_11000.txt.gz', chip_vec)
name_vec = gsub('(.*)_fcoc_hg19.*', '\\1', chip_vec)

tss_import_list = lapply(paste0('../raw_data/', tss_file_vec), read.table,
                         sep='\t', skip=1, stringsAsFactors=T, row.names=4)
tes_import_list = lapply(paste0('../raw_data/', tes_file_vec), read.table,
                         sep='\t', skip=1, stringsAsFactors=T, row.names=4)
names(tss_import_list) = names(tes_import_list) = name_vec
```


## General Transcription Factors around TSS

There might be differences in the occupancy of general transcription factors around
escapers, since they have to deal with different barriers from normal promoters.


```{r, fig.width=10, fig.height=15}

tss_mask_list = lapply(tss_import_list, run_mask, p_subset, start=-5500,
                       end=5500, step=20)
tss_mean_list = lapply(name_vec, run_group_mean, tss_mask_list, p_subset, start=-5500,
                       end=5500, step=20)



class_data = do.call(rbind.data.frame, tss_mean_list)


ggplot(class_data, aes(x=pos, y=mean, color=class)) +
    geom_line() +
    ggtitle('histone modifications around transcription start site') +
    scale_color_manual(values=COL_class) +
    facet_wrap(~name, nrow=4)

```



## General Transcription Factors around TES


```{r, fig.width=10, fig.height=15}

tes_mask_list = lapply(tes_import_list, run_mask, p_subset, start=-5500,
                       end=5500, step=20)
tes_mean_list = lapply(name_vec, run_group_mean, tes_mask_list, p_subset, start=-5500,
                       end=5500, step=20)


class_data = do.call(rbind.data.frame, tes_mean_list)

ggplot(class_data, aes(x=pos, y=mean, color=class)) +
    geom_line() +
    ggtitle('histone modifications around transcription termination site') +
    scale_color_manual(values=COL_class) +
    facet_wrap(~name, nrow=4)


```


```{r, fig.width=10, fig.height=30}

tssr_file_vec = gsub('.bigWig', '_tssr.txt.gz', chip_vec)
body_file_vec = gsub('.bigWig', '_body.txt.gz', chip_vec)
tssr_import_list = lapply(paste0('../raw_data/', tssr_file_vec), read.table,
                          row.names=4, stringsAsFactors=F,
                          col.names=c('chrom', 'start', 'end', 'name', 'score',
                                     'strand', 'size', 'num_data', 'min', 'max',
                                     'mean', 'sum'))
body_import_list = lapply(paste0('../raw_data/', body_file_vec), read.table,
                          row.names=4, stringsAsFactors=F,
                          col.names=c('chrom', 'start', 'end', 'name', 'score',
                                      'strand', 'size', 'num_data', 'min', 'max',
                                      'mean', 'sum'))

names(tssr_import_list) = names(body_import_list) = name_vec

for (name in name_vec){
    tssr_import_list[[name]]$mean = tssr_import_list[[name]]$mean /
                                    peak_import_list[[name]]$mean
    body_import_list[[name]]$mean = body_import_list[[name]]$mean /
                                    peak_import_list[[name]]$mean
}

tssr_list = lapply(tssr_import_list,
                   function(tssr, P){
                       data.frame(mean=tssr[P$name, 'mean'], class=P$class_GROcap,
                                  region='tss region (-50bp:+300bp)',
                                  log10=pseudo_log10(tssr[P$name, 'mean']))
                   }, P=p_subset)

body_list = lapply(body_import_list,
                   function(body, P){
                       data.frame(mean=body[P$name, 'mean'], class=P$class_GROcap,
                                  region='gene body (+3000bp:+20000bp)',
                                  log10=pseudo_log10(body[P$name, 'mean']))
                   }, P=p_subset)

names(tssr_list) = names(body_list) = name_vec

data_list = lapply(name_vec, function(name){
                       data = rbind.data.frame(tssr_list[[name]],
                                               body_list[[name]])
                       data$chip_name = name
                       return(data)
                   })

data = do.call(rbind.data.frame, data_list)

ggplot(data, aes(y=log10, x=class, color=class)) +
    geom_violin() +
    ggtitle(name) +
    geom_point(alpha=0.4, size=0.4,
               position=position_jitter(width=0.1)) +
    facet_wrap(~chip_name + region, ncol=4) +
    scale_color_manual(values=COL_class)
```
